---
title: "Advanced R Coding"
author: "BIOS 635"
date: "4/1/2021"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(caret)
library(rpart)
library(randomForest)
library(e1071)
library(DMwR)
```

```{r}
cancer_data_reg <- read_csv("../data/cancer_reg.csv") %>%
  select(-avgAnnCount, -avgDeathsPerYear, -incidenceRate, -Geography, -binnedInc) %>%
  drop_na()

heart_disease_data <- read_csv("../data/heart_disease/Correct_Dataset.csv",
                               na = c("", "NA", "?"))  %>%
  mutate(heart_disease = 
           relevel(factor(ifelse(Target>0, "Yes", "No")), 
                   ref = "No")) %>%
  select(-X1, -Target) %>%
  drop_na()

set.seed(12)
fetal_health_data <- read.csv("../data/fetal_health_subset.csv") %>% 
  select(baseline.value, accelerations, fetal_movement,
         uterine_contractions, light_decelerations, severe_decelerations,
         prolongued_decelerations, abnormal_short_term_variability,
         mean_value_of_short_term_variability, mean_value_of_long_term_variability,
         fetal_health) %>%
 drop_na() %>%
 mutate(fetal_health_cat = 
          factor(fetal_health)) %>%
  select(-fetal_health)
```

# Indexing objects
- Vectors
  - Created by `c(...)`
  - Indexed by `[]`
  - **All values must be same type**
- Matrices
  - Created by `matrix(...)`
  - Indexed by `[,]`
  - **All values must be same type**
- Lists
  - Created by `list()`
  - Indexed by `[]` or `[[]]` or `x$...`
  - Elements **can be of arbitrary types**
- Data frame
  - Created by `data.frame()`
  - Are **actually just lists**

```{r, error=TRUE}
x <- c(1,2,3)
x[1]

x <- list(c(1,2,3), c(4,5,6))
x[1]
x[[1]]

typeof(cancer_data_reg)
cancer_data_reg[1]
cancer_data_reg[[1]]
cancer_data_reg$TARGET_deathRate

# Objects can be dynamically expanded inside loops
x <- list()
y <- c()

for(i in 1:10){
  x[[i]] <- i
  y[i] <- i
}

# Can also name elements
names(x)
names(x) <- c("a", "b", "c")
x$a
```

# Nested Loops
- Can include loops **within loops**
- Various ways to keep track of indices, for example **creating a counter**

```{r, error=TRUE}
x <- list()
for(i in 1:10){
  print(i)
}

for(i in 1:10){
  paste0("i = ", i)
  for(j in 1:10){
    paste0("j = ", j)
  }
}

counter <- 1

for(i in 1:10){
  paste0("i = ", i)
  for(j in 1:10){
    paste0("j = ", j)
    x[[counter]] <- i+j
    
    counter <- counter+1
  }
}
```

# Functional Programming
- Functions are procedures programmed in R which dynamically run based on chosen input
- Have been working with functions from start of class (`train` for example)
- We can create our own functions to avoid repetitive code

```{r, error=TRUE}
# Suppose we want to run a KNN fitting method where we train, tune, test, and saved the results
rmse <- list()
tt_indices <- createFolds(y=cancer_data_reg$TARGET_deathRate, k=5)
for(i in 1:length(tt_indices)){
  cancer_train <- cancer_data_reg[-tt_indices[[i]],]
  cancer_test <- cancer_data_reg[tt_indices[[i]],]
  
  # Train and tune KNN
  algo_fit <- train(TARGET_deathRate~., data = cancer_train, preProcess = c("center", "scale"),
                    tuneLength = 10, trControl = trainControl(method="cv", number=5), method="knn")
  
  # Test on fold
  test_predict <- predict(algo_fit, newdata = cancer_test)
  rmse[[i]] <- postResample(pred = test_predict, obs = cancer_test$TARGET_deathRate)
}

all_results <- do.call("rbind", rmse)
cv_estimate <- apply(all_results, 2, mean)
cv_se <- apply(all_results, 2, sd)
final_results <- c("cv_rmse"=cv_estimate, "cv_se"=cv_se)
final_results

# What if we want to change the method?  Need to rewrite all of this code and just one thing.
# Use a function instead

pred_analysis <- function(data, outcome, k_train, k_tune, method, seed=12){
  rmse <- list()
  set.seed(seed)
  tt_indices <- createFolds(y=data[[outcome]], k=k_train)
  for(i in 1:length(tt_indices)){
    data_train <- data[-tt_indices[[i]],]
    data_test <- data[tt_indices[[i]],]
    
    # Train and tune KNN
    algo_fit <- train(as.formula(paste0(outcome, "~.")), data = data_train, preProcess = c("center", "scale"),
                      tuneLength = 10, trControl = trainControl(method="cv", number=k_tune), method=method)
    
    # Test on fold
    test_predict <- predict(algo_fit, newdata = data_test)
    rmse[[i]] <- postResample(pred = test_predict, obs = data_test[[outcome]])
  }
  
  all_results <- do.call("rbind", rmse)
  cv_estimate <- apply(all_results, 2, mean)
  cv_se <- apply(all_results, 2, sd)
  final_results <- c("cv_rmse"=cv_estimate, "cv_se"=cv_se)
  final_results <- list(c("cv_rmse"=cv_estimate, "cv_se"=cv_se), bestTune = algo_fit$bestTune)
  return(final_results)
}

# Run analysis
pred_analysis(data=cancer_data_reg, outcome="TARGET_deathRate", k_train = 5, k_tune = 5, method="knn")

# How about CART
pred_analysis(data=cancer_data_reg, outcome="TARGET_deathRate", k_train = 5, k_tune = 5, method="rpart")

# LASSO?
pred_analysis(data=cancer_data_reg, outcome="TARGET_deathRate", k_train = 5, k_tune = 5, method="glmnet")
# Be careful about tuning parameters!
```

# Apply functions
- Notice use of `apply` above
- Class of *apply* functions can be used to easy run desired function on many inputs
- Main examples: `apply` and `lapply`

```{r}
# Apply will run desired function on all elements of an array (rows or columns)
# By columns
apply(cancer_data_reg, MARGIN=2, FUN=mean)
apply(cancer_data_reg, MARGIN=2, FUN=sd)

# By rows
# apply(cancer_data_reg, MARGIN=1, FUN=mean)

# Can also specify function arguments in ... (for example na.rm=TRUE)
cancer_data_reg <- read_csv("../data/cancer_reg.csv") %>%
  select(-avgAnnCount, -avgDeathsPerYear, -incidenceRate, -Geography, -binnedInc)

apply(cancer_data_reg, MARGIN=2, FUN=mean)
apply(cancer_data_reg, MARGIN=2, FUN=mean, na.rm=TRUE)

# lapply is very similar, but will apply function to each element of a list
# Output returned in list form
lapply(cancer_data_reg, FUN=mean, na.rm=TRUE)
# Recall data.frames are lists, which each column being a different list element
# Hence, can use either apply or lapply

```
