---
title: "Summarize Results"
author: "Chris Moore"
date: "5/1/2021"
output:
  html_document:
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 10, fig.height = 5, message=FALSE) 
```
<br/><br/>
***NOTE: Click 'Code' on the right to expand code sections.***    
<br/><br/>

Packages   

```{r packages, results='hold', message=FALSE}
library(tidyverse)
library(data.table) 
library(gtsummary)
library(caret)
library(glmnet)
library(splines)
library(flextable)
```
<br/>   


Custom functions for formatting    

```{r custom_functions}
fmt_meanse <- function(x, dig=3, fmt = "f", flg="#") {
  paste0(formatC(mean(x), digits=dig, format = fmt, flag=flg),          # mean
         " (", formatC(sd(x), digits=dig, format = fmt, flag=flg), ")") #(SE)
}

factr_lvls <- function(dat) {
  dat %>% mutate(
    Method = factor(mod, levels=c("lso", "lso.sp", "rf"),
                    labels=c("LASSO", "LASSO (sp)", "Random Forest")),
    Event = factor(obs, levels=c("1", "0"), labels=c("Yes", "No")),
    stat = factor(obs, levels=c("1", "0"), labels=c("Sensitivity", "Specificity"))
  ) %>% select(Method, Event, stat, everything(), mod, obs)
}

lab_lso.sp <- as_paragraph("LASSO", as_sub("sp"))             # For flextables 
lab_plt <- c("LASSO", bquote("LASSO"["sp"]), "Random Forest") # For plots
```

<br/>  


Variable labels    
```{r variable_labels}
vrbls <- rbind(AGE_YRS = "Age (years)",
               SEX = "Sex",
               VAX_DOSE_SERIES = "Second vaccine dose",
               VAX_MANU = "Vaccine manufacturer",
               PriorAllergy = "Prior allergy", 
               PriorVaxAE = "Prior vaccine adverse event", 
               cur_htn = "Prevalent HTN", 
               cur_dm = "Prevalent diabetes", 
               cur_copd = "Prevalent COPD", 
               cur_cancer = "Prevalent cancer", 
               cur_hf = "Prevalent heart failure", 
               cur_MI = "History of MI", 
               cur_ckd = "Prevalent CKD", 
               cur_dialysis = "On dialysis", 
               cur_stroke = "History of stroke") %>% 
  as.data.frame() %>% rownames_to_column("v") %>% rename(lab = V1)
```

**********************    
<br/> <br/>  



## Import data and summarize performance

```{r import}
load("./Outputs/kfold_output/kfold_output_RF.RData")
load("./Outputs/kfold_output/kfold_output_LogLasso.RData")
load("./Outputs/kfold_output/kfold_output_LogLasso_se0.8sp1.2.RData")
```


Separate components of `kfold_cv()` custom function

```{r kfold_components}
m_rf <- rf_results$mod_train
p_rf <- rf_results$preds_test
e_rf <- rf_results$perf_test

m_lso <- loglso_results$mod_train
p_lso <- loglso_results$preds_test
e_lso <- loglso_results$perf_test

m_lso.sp <- loglso_results_wtd$mod_train
p_lso.sp <- loglso_results_wtd$preds_test
e_lso.sp <- loglso_results_wtd$perf_test
```


Combine fold-specific errors and summarize performance by class (sensitivity and specificity)

```{r get_classerr}
e_all <- rbindlist(list(lso=e_lso, lso.sp=e_lso.sp, rf=e_rf), idcol = "mod") %>% 
  factr_lvls()
  
sens_spec <- e_all %>% 
  group_by(mod, obs) %>% 
  summarise_at(vars(p_corr=Accuracy, p_incor=Error), 
               list(mn = ~mean(.), se = ~sd(.),
                    mnse = ~fmt_meanse(.),
                    p_mnse = ~fmt_meanse(.*100, dig=1))) %>%
  ungroup() %>% 
  factr_lvls() 
```


Use predictions to also get 2x2 table cell counts

```{r get_counts}
cell_cnts <- rbindlist(list(lso = p_lso[, .(f, obs, pred, correct)],
                            lso.sp = p_lso.sp[, .(f, obs, pred, correct)],
                            rf = p_rf), idcol = "mod") %>% 
  .[, .(.N, n_corr=sum(correct), n_incor = .N-sum(correct)), 
    by=c("mod","obs")]
```


Use predictions to get overall accuracy  

```{r get_acc}
ovral_acc <- rbindlist(list(lso = p_lso[, .(f, obs, pred, correct)], 
                            lso.sp = p_lso.sp[, .(f, obs, pred, correct)],
                            rf = p_rf), idcol = "mod") %>% 
  .[, .(acc=sum(correct)/.N), by=c("mod", "f")] %>% 
  group_by(mod) %>% 
  summarise(ovrl_acc = mean(acc), se_ovrlacc = sd(acc), 
            Accuracy = fmt_meanse(acc), p_Accuracy = fmt_meanse(acc*100, dig=1))
```


Combine performance summaries 

```{r perf_summary}
summ_perf <- sens_spec %>% 
  left_join(cell_cnts, by=c("mod", "obs")) %>% 
  left_join(ovral_acc, by=c("mod"))
```

**************************************
<br/>  <br/>  



## Tables

**Table 2:** Sensitivity, specificity, and accuracy   

```{r tab_2}
tab2 <- summ_perf %>% 
  select(Method, stat, p_corr_mnse, Accuracy) %>% 
  pivot_wider(names_from = "stat", values_from = "p_corr_mnse")

(tab2_flx <- flextable(tab2) %>% autofit() %>% 
    compose(i=2, j=1, value=lab_lso.sp))
```

<br/>  <br/>  



**Table 3:** Sensitivity, specificity, and accuracy   

```{r tab_3}
tab3 <- summ_perf %>% 
  select(Method, Event, N, 
         "N Correct"=n_corr, 
         "N Incorrect"=n_incor, 
         "% Correct (SE)" = p_corr_p_mnse,
         "% Overall accuracy (SE)"= p_Accuracy) 

(tab3_flx <- flextable(tab3) %>% autofit() %>% 
    compose(i=3:4, j=1, value=lab_lso.sp) %>% 
    align_text_col(align = "right", header = TRUE, footer = TRUE) %>% 
    merge_v(j=7) %>% fix_border_issues(part = "all"))
```

<br/>  <br/>  



**Table 4:** Average and fold-specific variable importance in the random forest  

```{r tab_4}
# Get values in wide format ##$
var_imprt <- lapply(m_rf, function(x) {
  rownames_to_column(varImp(x)[["importance"]],"vrbl")
}) %>% 
  rbindlist(idcol="fold") 

tab4 <- var_imprt %>% 
  pivot_wider(names_from = "fold", values_from = "Overall") %>% 
  mutate(vrbl2 = case_when( grepl("SEX", vrbl) ~ "SEX",
                            grepl("VAX_DOSE", vrbl) ~ "VAX_DOSE_SERIES",
                            grepl("VAX_MANU", vrbl) ~ "VAX_MANU",
                            TRUE ~ gsub("1", "", vrbl)),
         vrbl2 = factor(vrbl2, levels=vrbls$v, label=vrbls$lab)) %>%
  group_by(Predictor = vrbl2) %>% 
  summarise_if(is.numeric, sum) %>% ungroup() %>% ungroup() %>% 
  rowwise() %>% 
  mutate(`Mean (SE)` = fmt_meanse(c_across(fold1:fold5), dig=2),
         avg = mean(c_across(fold1:fold5))) %>%
  rename_with(~ gsub("fold", "Fold ", .)) %>% 
  arrange(-avg) %>% select(Predictor, `Mean (SE)`, everything())

(tab4_flx <- flextable(select(tab4, -avg)) %>% 
  autofit())
```

<br/>  <br/>  




************************
<br/>  <br/>  

## Figures 


**Figure 1:** Mean (standard error) LASSO and random forest predictive performance during five-fold cross-validation  

```{r, fig.height=5, fig.width=7}
fig1 <- summ_perf %>% 
  select(Method, stat, mn=p_corr_mn, se=p_corr_se, 
         mn_Accuracy=ovrl_acc, se_Accuracy=se_ovrlacc) %>% 
  pivot_wider(names_from = "stat", values_from = c("mn", "se")) %>% 
  pivot_longer(cols=-Method) %>% 
  separate(name, into=c("mnse", "stat"), sep="_") %>% 
  pivot_wider(names_from = "mnse") %>% 
  ggplot(aes(x=stat, y=mn, color=Method)) +
  geom_point(size=3) +
  geom_errorbar(aes(ymin=mn-se, ymax=mn+se), size=1, width=0.2) +
  scale_y_continuous("Mean (SE)", limits=c(0, 1), breaks=seq(0, 0.8, 0.2), expand=c(0,0)) +
  xlab(NULL) +
  scale_color_brewer(palette = "Set1", labels=lab_plt) +
  theme_bw(base_size = 15, base_line_size=0.3)
fig1
```

<br/>  <br/>  



**Figure 2:**  LASSO predicted probabilities by observed event status. Dashed line represents average threshold selected for predicting an event from ROC analyses

```{r, fig.width = 10, fig.height = 6}
## Re-create fold data ##$##$
source("./R/setup_data.R")
set.seed(12) # same seed as used before
folds <- createFolds(y=unlist(dat[, event_asnum]), k=5)

d_fold <- lapply(folds, function(f) {
  dat[f]
}) %>% rbindlist(idcol="fold")

## Combine with predictions ##$##$
preds <- cbind(d_fold, p_lso)
preds[, Event := factor(obs, levels=c("0", "1"), labels=c("No", "Yes"))]
# Check that data lines up (observerd outcomes should match)
# sum(preds$SeriousEvent != preds$obs)
##$##$##$##$##$##$



# Average prob threshold from ROC ##$##$
p_thresh = data.frame(
  Analysis = c("LASSO", "LASSO (sp)"),
  Threshold = c(mean(sapply(m_lso, `[[`, "p_thresh")), mean(sapply(m_lso.sp, `[[`, "p_thresh")))
)
##$##$##$##$##$##$



## Plot prob densities by event ##$##$
p1 <- ggplot(preds, aes(x=pred_raw)) +
  geom_density(aes(y=..scaled.., fill=Event, group=Event), alpha=0.3) +
  geom_vline(data=p_thresh, aes(xintercept=Threshold, color=Analysis), 
             size=0.72, linetype="dashed") +
  scale_y_continuous("Density", limits = c(0, 1.1),
                     breaks = seq(0, 1, 0.2), expand=c(0,0)) +
  scale_x_continuous("Predicted probability of event", limits = c(0, 1.1),
                     breaks = seq(0, 1, 0.2), expand=c(0,0)) +
  scale_fill_brewer(palette = "Set1") +
  scale_color_manual(values=c("darkgreen", "purple"), labels=lab_plt[1:2]) +
  
  guides(linetype = guide_legend(override.aes = list(size = 0.9))) +
  theme_bw(base_size = 16, base_line_size=0) +
  theme(legend.box.margin = margin(t=-5, b=-7),
        legend.key.height = unit(1.5, "line"),
        legend.title = element_text(face="bold"))

## Plot probs vs age by event ##$##$
p2 <- ggplot(preds[AGE_YRS<quantile(AGE_YRS,0.99)], # Trim at 99th percentile
             aes(y=pred_raw, x=AGE_YRS, color=Event)) +
  geom_point(alpha=0.13, size=0.6) +
  geom_hline(data=p_thresh, aes(yintercept=Threshold, color=Analysis), size=0.76, linetype="dashed") +
  facet_wrap(.~Event, ncol=1) +
  scale_x_continuous("Age (years)", limits = c(17, 92),
                     breaks = seq(20, 90, 20), expand=c(0,0)) +
  scale_y_continuous("Predicted probability of event", limits = c(0, 1.1),
                     breaks = seq(0, 1, 0.25), expand=c(0,0)) +
  scale_color_manual(values=c("darkgreen", "purple", "red", "blue")) +
  # scale_color_brewer(palette = "Set1") +
  scale_linetype_manual(values=c("dashed", "twodash")) +
  theme_bw(base_size = 16, base_line_size=0.3) +
  theme(strip.text = element_blank(),
        strip.background = element_blank())
##$##$##$##$##$##$



## Combine ##$##$
fig2 <- ggpubr::ggarrange(p1, p2, ncol=2, common.legend = TRUE)
fig2
```
<br/>  <br/>  



**Figure 3** Random forest number of predictors selected for each node versus out-of-bag error   

```{r fig.width = 10, fig.height = 6}
npreds_rf <-  lapply(m_rf, `[[`, "results") %>% 
  rbindlist(idcol="Fold") %>%
  mutate(Fold = gsub("fold", "", Fold))

fig3 <- npreds_rf %>% 
  mutate(se=NA, categ="By fold") %>% 
  rbind(npreds_rf %>% 
          group_by(mtry) %>% 
          summarise(categ = "Mean (SE)", Fold="Mean", se = sd(Accuracy), 
                    Accuracy = mean(Accuracy), Kappa = mean(Kappa))) %>% 
  ggplot(aes(x=mtry, y=Accuracy, color=Fold)) +
  geom_point(size=3) +
  geom_line(size=1) +
  geom_errorbar(aes(ymin=Accuracy-se, ymax=Accuracy+se), size=1, width=0.5) +
  facet_grid(. ~ categ) +
  scale_y_continuous("OOB Accuracy", limits = c(0.85, 0.90), 
                     breaks = seq(0.85, 0.9, 0.01), expand=c(0,0)) +
  scale_x_continuous("Number of predictors", limits = c(3, 17),
                     breaks = seq(4, 16, 2), expand=c(0,0)) +
  scale_color_manual(values=c("red", "blue", "green", "purple", "orange", "black")) +
  theme_bw(base_size = 16, base_line_size=0.3)
fig3
```

<br/>  <br/>  



**Figure 4:** Random forest variable importance averaged over folds   

```{r}
fig4 <- ggplot(tab4, aes(x = Predictor, y=avg)) +
  geom_bar(stat="identity") +
  scale_y_continuous("Average variable importance", limits = c(0, 101),
                     breaks = seq(0, 100, 20), expand=c(0,0)) +
  scale_x_discrete(limits=rev) +
  scale_color_brewer(palette = "Set1") +
  coord_flip() +
  theme_bw(base_size = 14, base_line_size=0.3)
fig4
```

**************************************
<br/>  <br/>  



## Save
```{r, eval=FALSE}
fpath <- "./Outputs/Final_Results/"

# Tables ##$
tbls <- paste0("tab", 2:4)
walk(paste0(tbls, "_flx"), ~ save_as_docx(get(.x), path=paste0(fpath, .x, ".docx")))
walk(tbls, ~ fwrite(get(.x), paste0(fpath, .x, ".csv")))

# Figures ##$
figs <- paste0("fig", 1:4)# ht x width      ht x width
fig_params <- list(fig1 = c(NA,NA), fig2 = c(10, 6),
                   fig3 = c(10, 6), fig4 = c(10.5,6))
walk(figs, ~ ggsave(
  plot=get(.x), filename=paste0(fpath, .x, ".jpeg"), units="in",
  width = fig_params[[.x]][[1]], height = fig_params[[.x]][[2]]
  ) )
```

**************************************
<br/>  <br/>  



## Appendix

These additional results about the final models are only for reference and are not formatted for presentation

#### LASSO models
```{r lasso_coeff}
lasso_coef <- lapply(m_lso, function(x) {
  tmp <- coef(x, s = "lambda.min")
  data.frame(name = tmp@Dimnames[[1]][tmp@i + 1], 
             coefficient = tmp@x)
} )


lasso_models <- data.frame(
  f = names(m_lso),
  lambda = sapply(m_lso, `[[`, "lambda.min"),
  n_terms = c(sapply(lasso_coef, nrow)),
  age_frmla = as.character(sapply(m_lso, `[[`, "cust_tune")),
  p_thresh = sapply(m_lso, `[[`, "p_thresh"),
  p_thresh.sp =  sapply(m_lso.sp, `[[`, "p_thresh"),
  row.names = NULL
) %>% mutate(age_frmla = gsub(")\\^2", "", gsub(".*\\+", "", age_frmla)))

lasso_models
```
<br/>  

#### Random forest tune parameters

```{r}
rf_tuneparams <- lapply(m_rf, function(x) {
  data.frame(n_tree = x[["finalModel"]][["ntree"]],
             n_vrbls = x[["finalModel"]][["mtry"]])
}) %>% rbindlist(idcol="fold")

rf_tuneparams
```
