---
title: "635 Final Project (Eileen's part)"
author: "Eileen Yang"
date: "4/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rlang)
library(reshape)
library(reshape2)
library(caret)
library(factoextra)
library(gtsummary)
library(flextable)
library(pROC)
library(glmnet)
library(broom)
library(gt)
library(caret)
library(GGally)
library(mgcv)
library(splines)
```

Read in data
```{r}

food <-  read_csv("/Users/eileenyang/Desktop/R/BIOS 635/final-project-group-6-1-main/data/Food_Supply_kcal_Data.csv") %>%
  #first removing variables that aren't of interest (ie. not relevant, not a food group, or are unlikely to have an real impact)
  select(-Recovered, -Active, -`Unit (all except Population)`, -Obesity, -Undernourished, -Miscellaneous, -Spices, -Country) %>%
  drop_na() %>%
  #cases are per 100,000,000 people
  mutate(Cases_per_capita = Confirmed/Population * 100000000, Deaths_per_capita = Deaths/Population * 100000000) %>%
  #creating a column that specifies whether each country falls above or below the median
  mutate(Cases_vs_median = ifelse(Cases_per_capita > median(Cases_per_capita), "Above", "Below"), Deaths_vs_median = ifelse(Deaths_per_capita > median(Deaths_per_capita), "Above", "Below"))

food$Cases_vs_median <- as.factor(food$Cases_vs_median)
food$Deaths_vs_median <- as.factor(food$Deaths_vs_median)

food <- select(food,  -c("Confirmed","Deaths", "Population"))

food_cases <- select(food,  -c("Deaths_per_capita","Deaths_vs_median"))
food_deaths <- select(food,  -c("Cases_per_capita","Cases_vs_median"))

```

Summarize data
```{r}
#remove country for data summary
foodcols <- colnames(food)
foodcols <- foodcols[-1]

#summary for cases_vs_median:
tbl1 <- tbl_summary(data=food_cases, by=Cases_vs_median, include=colnames(food_cases), statistic = list(all_continuous() ~ "{mean} ({sd})"))

tbl2 <- tbl1 %>% add_p(test = list(all_continuous() ~ "aov")) %>% add_n()
tbl2_flex <- as_flex_table(tbl2)  %>% bold(part="header")
tbl2_flex

#summary for cases and deaths_vs_median:
tbl1 <- tbl_summary(data=food_deaths, by=Deaths_vs_median, include=colnames(food_deaths), statistic = list(all_continuous() ~ "{mean} ({sd})"))

tbl2 <- tbl1 %>% add_p(test = list(all_continuous() ~ "aov")) %>% add_n()
tbl2_flex <- as_flex_table(tbl2)  %>% bold(part="header")
tbl2_flex

#######################

#summary for cases per capita:
tbl1 <- tbl_summary(data=food, include=colnames(food), statistic = list(all_continuous() ~ "{mean} ({sd})"))

tbl2 <- tbl1  %>% add_n()
tbl2_flex <- as_flex_table(tbl2)  %>% bold(part="header")
tbl2_flex

```

## Correlations
```{r, fig.width = 12, fig.height = 8}
#First, some correlation plots to see correlations
food_cases_num <- food_cases[,-23]
foodcases_corr <- cor(food_cases_num)
corrp <- corrplot::corrplot(foodcases_corr, method="number")
```
- It looks like some food categories are strongly associated with others:
  - Milk - Excluding Butter and Animal Products (+)
  - Meat and Animal Products (-)
  - Vegetal Products and Animal Products (-)
  - Vegetal Products and Animal Fats (-)
  - Vegetal Products and Meat (-)
  - Vegetal Products and Milk - Excluding Butter (-)

- It would make sense to remove some of these variables--perhaps removing Animal Products and Vegetal Products would be beneficial. **(or even more of the animal-related variables?)**


# Continuous Outcomes - Cases
**note I had to remove Aquatic Products, Other for now bc it was giving me some issues**
With collinear variables removed (Animal Products, Vegetal Products)
```{r}
food_cases2 <- select(food_cases, -c("Cases_vs_median","Aquatic Products, Other", "Animal Products","Vegetal Products"))
```

## KNN regression (10 fold)
```{r}
set.seed(12)

  folds <- createFolds(y=food_cases2$Cases_per_capita, k=10)
  foldnum <- list()
  mselist <- list()
  
  for(f in 1:length(folds)){
    foldnum[f] <- paste("Fold", f)
    foodcases_train <- food_cases2[-folds[[f]],]
    foodcases_test <- food_cases2[folds[[f]],]
    knn_train <- train(form=`Cases_per_capita` ~., data=foodcases_train,
                       method="knn", tuneLength = 20, preProcess=c("scale", "center"))
    foodcases_pred <- predict(knn_train, newdata=foodcases_test)
    mselist[f] <- mean((foodcases_test$`Cases_per_capita` - foodcases_pred) ^ 2) 
    
    }
error_knn <- mean(unlist(mselist))
sd_knn <- sd(unlist(mselist))
error_knn
sd_knn

```

## Linear Regression

First check assumptions 
**Actually doesn't seem to meet lin regression assumptions...so might take this part out**
```{r, fig.width = 12, fig.height = 8}
food_cases2 %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
    facet_wrap(~ key, scales = "free") +
    geom_histogram()


#removing outliers
#length(which(food_cases2$Cases_per_capita > 500))
#mean(food_cases2$Cases_per_capita)
#hist(food_cases2[food_cases2$Cases_per_capita < 500,]$`Cases_per_capita`)

food_cases2_outlier_rm <- food_cases2[food_cases2$Cases_per_capita < 500,]


#melt your data
foodcases_melt <- melt(food_cases2_outlier_rm,"Cases_per_capita")
head(foodcases_melt)


#scatterplot per group
ggplot(foodcases_melt, aes(x=value, y=Cases_per_capita)) +
  geom_point() +
  facet_wrap(.~variable, scales = "free_x") 
```

10 fold CV
```{r}
set.seed(12)

folds <- createFolds(y=food_cases2$Cases_per_capita, k=10)
  foldnum <- list()
  mselist2 <- list()
  
for(f in 1:length(folds)){
    foldnum[f] <- paste("Fold", f)
    foodcases_train <- food_cases2[-folds[[f]],]
    foodcases_test <- food_cases2[folds[[f]],]
    lm_train <- lm(`Cases_per_capita` ~., data=foodcases_train)
    foodcases_pred <- predict(lm_train, newdata=foodcases_test)
    mselist2[f] <- mean((foodcases_test$`Cases_per_capita` - foodcases_pred) ^ 2)
}
error_lin <- mean(unlist(mselist2))
sd_lin <- sd(unlist(mselist2))
error_lin
sd_lin
```


## Lasso vs linear regression - holdout method to see which variables would be removed
```{r}
food_cases2_forlasso <- select(food_cases, -c("Cases_vs_median","Aquatic Products, Other"))

set.seed(12)

#From Lasso
foodcases_lasso_full <- cv.glmnet(x = as.matrix(food_cases2_forlasso[, -21]), y = food_cases2$Cases_per_capita, alpha = 1) 
tmp_coeffs <- coef(foodcases_lasso_full, s=foodcases_lasso_full$lambda.min)
data.frame(name = tmp_coeffs@Dimnames[[1]][tmp_coeffs@i + 1], coefficient = tmp_coeffs@x)
plot(foodcases_lasso_full)

#From Linear Regression
foodcases_lm_full <- lm(Cases_per_capita ~., data=food_cases2_forlasso)
summary(foodcases_lm_full)
```
- It looks like 12 variables are kept in the model, with 8 removed: Alcoholic Beverages, Animal Products, Cereals - excluding Beer, Milk - Excluding Butter, Oilcrops, Sugar Crops, Treenuts, and Vegetal Products.


## Lasso (10-fold):
```{r}
set.seed(12)
folds <- createFolds(y=food_cases2_forlasso$Cases_per_capita, k=10)
  foldnum <- list()
  mselist2 <- list()
  
  for(f in 1:length(folds)){
    foldnum[f] <- paste("Fold", f)
    foodcases_train <- food_cases2_forlasso[-folds[[f]],]
    foodcases_test <- food_cases2_forlasso[folds[[f]],]

    foodcases_lasso1 <- cv.glmnet(x=as.matrix(foodcases_train[,-21]), y=foodcases_train$Cases_per_capita, alpha=1)
    opt_lambda <- foodcases_lasso1$lambda.min
    
    foodcases_pred <-  predict(foodcases_lasso1, newx=as.matrix(foodcases_test[,-21]), s=opt_lambda)
    mselist2[f] <- mean((foodcases_test$Cases_per_capita - foodcases_pred) ^ 2) 
    
  }
error_lasso <- mean(unlist(mselist2))
sd_lasso <- sd(unlist(mselist2))

error_lasso
sd_lasso

```

Ridge - 10 fold
```{r}
set.seed(12)
folds <- createFolds(y=food_cases2_forlasso$Cases_per_capita, k=10)
  foldnum <- list()
  mselist2 <- list()
  
  for(f in 1:length(folds)){
    foldnum[f] <- paste("Fold", f)
    foodcases_train <- food_cases2_forlasso[-folds[[f]],]
    foodcases_test <- food_cases2_forlasso[folds[[f]],]

    foodcases_ridge1 <- cv.glmnet(x=as.matrix(foodcases_train[,-21]), y=foodcases_train$Cases_per_capita, alpha=0)
    opt_lambda <- foodcases_ridge1$lambda.min
    
    foodcases_pred <-  predict(foodcases_ridge1, newx=as.matrix(foodcases_test[,-21]), s=opt_lambda)
    mselist2[f] <- mean((foodcases_test$Cases_per_capita - foodcases_pred) ^ 2) 
    
  }
error_ridge <- mean(unlist(mselist2))
sd_ridge <- sd(unlist(mselist2))

error_ridge
sd_ridge
```

***

# Continuous Outcomes - Deaths
**note I had to remove Aquatic Products, Other for now bc it was giving me some issues**
With collinear variables removed (Animal Products, Vegetal Products)
```{r}
food_deaths2 <- select(food_deaths, -c("Deaths_vs_median","Aquatic Products, Other", "Animal Products","Vegetal Products"))
```

## KNN regression (10 fold)
```{r}
set.seed(12)

  folds <- createFolds(y=food_deaths2$Deaths_per_capita, k=10)
  foldnum <- list()
  mselist <- list()
  
  for(f in 1:length(folds)){
    foldnum[f] <- paste("Fold", f)
    fooddeaths_train <- food_deaths2[-folds[[f]],]
    fooddeaths_test <- food_deaths2[folds[[f]],]
    knn_train <- train(form=`Deaths_per_capita` ~., data=fooddeaths_train,
                       method="knn", tuneLength = 20, preProcess=c("scale", "center"))
    fooddeaths_pred <- predict(knn_train, newdata=fooddeaths_test)
    mselist[f] <- mean((fooddeaths_test$`Deaths_per_capita` - fooddeaths_pred) ^ 2) 
    
    }
error_knn <- mean(unlist(mselist))
sd_knn <- sd(unlist(mselist))
error_knn
sd_knn

```

## Linear Regression

First check assumptions 
**Actually doesn't seem to meet lin regression assumptions...so might take this part out**
```{r, fig.width = 12, fig.height = 8}
food_deaths2 %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
    facet_wrap(~ key, scales = "free") +
    geom_histogram()


#removing outliers
#length(which(food_deaths2$Deaths_per_capita > 15))
#median(food_deaths2$Deaths_per_capita)

food_deaths2_outlier_rm <- food_deaths2[food_deaths2$Deaths_per_capita < 15,]


#melt your data
fooddeaths_melt <- melt(food_deaths2_outlier_rm,"Deaths_per_capita")
head(fooddeaths_melt)


#scatterplot per group
ggplot(fooddeaths_melt, aes(x=value, y=Deaths_per_capita)) +
  geom_point() +
  facet_wrap(.~variable, scales = "free_x") 
```

10 fold CV
```{r}
set.seed(12)

folds <- createFolds(y=food_deaths2$Deaths_per_capita, k=10)
  foldnum <- list()
  mselist2 <- list()
  
for(f in 1:length(folds)){
    foldnum[f] <- paste("Fold", f)
    fooddeaths_train <- food_deaths2[-folds[[f]],]
    fooddeaths_test <- food_deaths2[folds[[f]],]
    lm_train <- lm(`Deaths_per_capita` ~., data=fooddeaths_train)
    fooddeaths_pred <- predict(lm_train, newdata=fooddeaths_test)
    mselist2[f] <- mean((fooddeaths_test$`Deaths_per_capita` - fooddeaths_pred) ^ 2)
}
error_lin <- mean(unlist(mselist2))
sd_lin <- sd(unlist(mselist2))
error_lin
sd_lin
```


## Lasso vs linear regression - holdout method to see which variables would be removed
```{r}
food_deaths2_forlasso <- select(food_deaths, -c("Deaths_vs_median","Aquatic Products, Other"))

set.seed(12)

#From Lasso
fooddeaths_lasso_full <- cv.glmnet(x = as.matrix(food_deaths2_forlasso[, -21]), y = food_deaths2$Deaths_per_capita, alpha = 1) 
tmp_coeffs <- coef(fooddeaths_lasso_full, s=fooddeaths_lasso_full$lambda.min)
data.frame(name = tmp_coeffs@Dimnames[[1]][tmp_coeffs@i + 1], coefficient = tmp_coeffs@x)
plot(fooddeaths_lasso_full)

#From Linear Regression
fooddeaths_lm_full <- lm(Deaths_per_capita ~., data=food_deaths2_forlasso)
summary(fooddeaths_lm_full)
```


## Lasso (10-fold):
```{r}
set.seed(12)
folds <- createFolds(y=food_deaths2_forlasso$Deaths_per_capita, k=10)
  foldnum <- list()
  mselist2 <- list()
  
  for(f in 1:length(folds)){
    foldnum[f] <- paste("Fold", f)
    fooddeaths_train <- food_deaths2_forlasso[-folds[[f]],]
    fooddeaths_test <- food_deaths2_forlasso[folds[[f]],]

    fooddeaths_lasso1 <- cv.glmnet(x=as.matrix(fooddeaths_train[,-21]), y=fooddeaths_train$Deaths_per_capita, alpha=1)
    opt_lambda <- fooddeaths_lasso1$lambda.min
    
    fooddeaths_pred <-  predict(fooddeaths_lasso1, newx=as.matrix(fooddeaths_test[,-21]), s=opt_lambda)
    mselist2[f] <- mean((fooddeaths_test$Deaths_per_capita - fooddeaths_pred) ^ 2) 
    
  }
error_lasso <- mean(unlist(mselist2))
sd_lasso <- sd(unlist(mselist2))

error_lasso
sd_lasso

```

Ridge - 10 fold
```{r}
set.seed(12)
folds <- createFolds(y=food_deaths2_forlasso$Deaths_per_capita, k=10)
  foldnum <- list()
  mselist2 <- list()
  
  for(f in 1:length(folds)){
    foldnum[f] <- paste("Fold", f)
    fooddeaths_train <- food_deaths2_forlasso[-folds[[f]],]
    fooddeaths_test <- food_deaths2_forlasso[folds[[f]],]

    fooddeaths_ridge1 <- cv.glmnet(x=as.matrix(fooddeaths_train[,-21]), y=fooddeaths_train$Deaths_per_capita, alpha=0)
    opt_lambda <- fooddeaths_ridge1$lambda.min
    
    fooddeaths_pred <-  predict(fooddeaths_ridge1, newx=as.matrix(fooddeaths_test[,-21]), s=opt_lambda)
    mselist2[f] <- mean((fooddeaths_test$Deaths_per_capita - fooddeaths_pred) ^ 2) 
    
  }
error_ridge <- mean(unlist(mselist2))
sd_ridge <- sd(unlist(mselist2))

error_ridge
sd_ridge
```
```

